---
title: "UWPCE-DS1-Asst2-JStearns.Rmd"
author: "Jim Stearns"
date: "18 October 2014"
fontsize: 14pt
---

Assignment 2, Introduction to Data Science, UW PCE, Fall 2014
====
```{r echo=FALSE}
#setwd("~/GoogleDrive/Learning/Courses/UWPCE-DataScience/Course1_Intro/Assignments")
```

> For all assignment items, use the patterns described in:
>
> * DataScience02a.R
> * DataScience02b.R
> * DataScience02c.R

1. Using R Data Preparation
====
a. Get Indian Liver Patient Dataset:
----
```{r}
url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/00225/Indian%20Liver%20Patient%20Dataset%20(ILPD).csv"
ILPD.df<-read.csv(url,header=FALSE,stringsAsFactors=FALSE)
str(ILPD.df)
```
b. Get the column headers
----
> From http://archive.ics.uci.edu/ml/datasets/ILPD+(Indian+Liver+Patient+Dataset)#

The link is an HTML page that contains a listing of the attributes.
The HTML is primitive - not even a table, just a paragraph with <br>eaks.

So I got the information manually. Copied and pasted, here it is:

Attribute Information:

1. Age  Age of the patient 
2. Gender	Gender of the patient 
3. TB	Total Bilirubin 
4. DB	Direct Bilirubin 
5. Alkphos Alkaline Phosphotase 
6. Sgpt Alamine Aminotransferase 
7. Sgot Aspartate Aminotransferase 
8. TP	Total Protiens 
9. ALB	Albumin 
10. A/G Ratio	Albumin and Globulin Ratio 
11. Selector field used to split the data into two sets (labeled by the experts) 

More detail on the 11th field, Selector: "Selector is a class label used to divide into groups(liver patient or not)."

c. Manually construct a vector of column names.
----
```{r}
headers <- c("Age", "Gender", "TB", "DB", "Alkphos", "Sgpt", "Sgot", "TP", "ALB", "A/G Ratio", "Selector") 
```
Most of the names provided by the study authors are valid R names.
But the slash and space in "A/G Ratio" look iffy. 
Use an R method that coerces names to syntactically valid names.
```{r}
valid_headers <- make.names(headers)
valid_headers
```
d. Associate names with the dataframe
----
```{r}
names(ILPD.df) <- valid_headers
str(ILPD.df)
```

2.Using R Data Exploration on ILPD
====
a. Use **head(ILPD)** to view the first six rows.
----
```{r}
head(ILPD.df, n=6)
```
b. Determine the mean, median, and standard deviation of each column
----
The Gender column is of type character ("Male", "Female"). Let's convert it to an integer
so that mean etc calculations can be performed on it.
```{r}
ILPD.df$Gender <- as.integer(as.factor(ILPD.df$Gender))
str(ILPD.df)
# R is 1-relative, so the values assigned here are 1 and 2. Make that 0 (Female) and 1 (Male)
ILPD.df$Gender <- ILPD.df$Gender - 1
str(ILPD.df)
```
First pass at calculations:
```{r}
sapply(ILPD.df, mean)
sapply(ILPD.df, median)
sapply(ILPD.df, sd)
```
A.G.Ratio returns a NA, which usually indicates at least one value is NA.
```{r}
print(sprintf("The number of NA values in A.G.Ratio = %d", sum(is.na(ILPD.df$A.G.Ratio))))
```
Finesse the NA problem by specifying NA values are to be removed before the mean et al 
are calculated:
```{r}
sapply(ILPD.df, mean, na.rm=TRUE)
sapply(ILPD.df, median, na.rm=TRUE)
sapply(ILPD.df, sd, na.rm=TRUE)
```
c. What does **na.rm=TRUE** do in sd(x, **na.rm=TRUE**)?
----
Skip rows that have a N/A value (applied above).

d. Create Histograms (**hist**) for each column where possible.
----
```{r}
str(ILPD.df)
par(mfrow=c(1,1))
for (col in seq(1, ncol(ILPD.df))) {
    xname <- names(ILPD.df)[col]
    hist(ILPD.df[,col], xlab=xname, main=paste("Histogram of", xname))
}
```

e. Use the **plot(ILPD)** function to present a general overview of the data.
----
```{r}
par(mfrow=c(1,1))
plot(ILPD.df)
# And plot to a PDF file so one can blow it up easily:
pdf("PlotILPD.pdf", width=7, height=7)
plot(ILPD.df)
dev.off()
```

> You want to see a matrix of many plots.
> Your efforts may be thwarted because the Gender column is not numeric.
> You can skip the Gender column, or you can turn the gender column into a numeric column.
> You might need help from a fellow student, the LinkedIn group, or me.
> Look at the plots and answer:

e-a. What can you say about the data?
----

* Three pairs of predictors show a strong correlation - noted below.
* While 75% of the subjects are male, despite this weighting of male subjects, men are more likely to have high concentrations of Sgpt (Alamine Aminotransferase).
* The plot of two bi-modal variables - Gender and Selection - yield 4 unhelpfully similar arcs in the four corners. Whether any correlation exists between gender and status as a liver patient requires further analysis.

e-b. How can you tell if a vector contains continuous numbers or binary data?
----
Binary data is bimodal. Continuous number are more distributed. Gender and Selector are bimodal, and binary.

e-c. How can you tell if two vectors are correlated?
----
On a 2D plot of the two vectors, are the observations roughly linear?

Looking at the plot output, one can see that TB:DB, TP:ALB and ALB:A.G.Ratio seem to be correlated.

3. Using Data Preparation Concepts
====

> Create examples of the following data preparation processes in R:

a. Remove Outliers
----
```{r}
testList <- c(-1, 1, 5, 1, 1, 17, -3, 1, 1, 3)
highLimit <- mean(testList) + 2*sd(testList)
lowLimit <- mean(testList) - 2*sd(testList)
goodFlag <- (testList < highLimit) & (testList > lowLimit)

print(sprintf("List with outliers removed: %s", paste(testList[goodFlag], collapse=", ")))
print(sprintf("Outliers removed: %s", paste(testList[!goodFlag], collapse=", ")))
```

b. Relabel 
----
```{r}
degree <- c('BS', 'MS', 'PhD', 'HS', 'Bachelors', 'Masters', 'High School', 'BS', 'MS', 'MS')
str(degree)
highSchoolDegrees=c("High School", "HS")
collegeDegrees=c("BS", "Bachelors")
graduateDegrees=c("MS", "PhD", "Masters")
degree[degree %in% highSchoolDegrees] <- "High School"
degree[degree %in% collegeDegrees] <- "College"
degree[degree %in% graduateDegrees ] <- "Graduate"
degree
```
c. Normalize: Min-Max and Z-score
----
```{r}
testVector <- c(-1, 1, 5, 1, 1, 17, -3, 1, 1, 3)

minMaxNormalize <- function(v) {
    minv <- min(v, na.rm=TRUE)
    maxv <- max(v, na.rm=TRUE)
    normv <- v - minv
    normv <- normv / (maxv - minv)
    return(normv)
}

zScoreNormalize <- function(v) {
    meanv <- mean(v, na.rm=TRUE)
    sdv <- sd(v, na.rm=TRUE)
    normv <- v - meanv
    normv <- normv / sdv
    return(normv)
}

testVectorMinMaxNormalized <- minMaxNormalize(testVector)
print(sprintf("Vector Min-Max normalized = %s", 
              paste(testVectorMinMaxNormalized, collapse=", ")))

testVectorZScoreNormalized <- zScoreNormalize(testVector)
print(sprintf("Vector Z-score normalized values (to two significant digits) = %s", 
               paste(format(testVectorZScoreNormalized, digits=1), collapse=", ")))
```
d. Binarize
----
```{r}
color <- c('Red', 'Green', 'Blue', 'Blue', 'Blue', 'Blue', 'Blue', 'Red', 'Green', 'Blue')

isRed <- color == 'Red' 
isGreen <- color == 'Green'
isBlue <- color == 'Blue'

isRed <- as.numeric(isRed)
isGreen <- as.numeric(isGreen)
isBlue <- as.numeric(isBlue)
isRed
isGreen
isBlue

# Make sure all vector elements have been accounted for - once
allThree <- isRed + isGreen + isBlue
allOnes <- rep(1, length(allThree))
allAccountedFor <- sum(allThree == allOnes) == length(allOnes)
print(sprintf("All vector elements accounted for - once == %s", allAccountedFor))
```
e. Discretize: three equal ranges, three equal amounts
----
```{r}
v <- c(3,4,4,5,5,5,5,5,5,5,5,6,6,6,6,6,7,7,7,7,8,8,9,12,23, 23, 25, 81)

NBINS = 3
range <- max(v) - min(v)
binWidth <- range / NBINS
bin2Min <- binWidth
bin3Min <- 2*binWidth

range
binWidth
bin2Min
bin3Min

vDiscretized <- rep(NA, length(v))
vDiscretized[v < bin2Min] <- "Low"
vDiscretized[v >= bin2Min & v < bin3Min] <- "Medium"
vDiscretized[v >= bin3Min] <- "High"
print(sprintf("Vector discretized into three bins: %s", paste(vDiscretized, collapse=", ")))
```